Measuring Directionality in Double Beta Decay and Neutrino Interactions with Kiloton-Scale Scintillation Detectors

### 2013-07-19 Direction.tex
Changes by Matt on the reconstruction section and Acknowledgments. Changes by Christoph and Lindley to sections 3 through 6. Addition of detector view with 0nuBB event (version 0). New references. Clean up of deprecated text and comments. July 19 (version 1 of draft). --Christoph

\section{Light Production in Liquid Scintillators}
Liquid scintillators are cocktails of aromatic hydrocarbons. When
charged particles move through a scintillator, the molecules are
excited, predominantly the the non-localized electrons in the
$\pi$-bonds of the phenyl groups~\cite{scintillator_ref}. Vibrational
and rotational modes of the molecules are turned into heat within
picoseconds through collisions with other molecules.  Within $\sim$10
picoseconds, the $\pi$-electrons de-excite to the first excited state
from higher levels through radiationless transitions. The first
excited state de-excites through photon emission. There are two
characteristic times for this de-excitation, depending if the singlet
state or the triplet state was excited.  The singlet state will
de-excite within nanoseconds while the triplet state de-excites on the
order of 10's or 100's of nanoseconds. These two processes are
fluorescence and phosphorescence respectively. The exact time
constants for these processes are determined by the composition of the
scintillator.

@@

A good liquid scintillator will produce $\sim$10,000 photons
isotropically per MeV of deposited energy. Although less abundant,
Cerenkov light will be produced as well if a particle is moving faster
than the speed of light in the medium.  This light is emitted in a
cone pointed in the direction of the particles trajectory and with a
continuous spectrum weighted toward shorter wavelengths
but extending well into the red. The spectrum is described by:
\begin{equation}
\label{eqCerenkov}
\frac{dN}{d\lambda dx} = \frac{2 \pi \alpha Z^2}{\lambda^2} ( 1 - \frac{1}{\beta^2 n(\lambda)^2} )
\end{equation}
where $n(\lambda)$ is the wavelength-dependent index of refraction and
$\beta$ is the velocity of the incoming particle. The Cerenkov light
produced at wavelengths shorter than the absorption cutoff of the
scintillator will be absorbed and re-emitted as isotropic light, but
wavelengths longer than this cutoff will propagate across the
detector, retaining their directional information. The yield is
roughly 60 photons per MeV, assuming a 400~nm absorption
cutoff\cite{qdot}. These undisturbed Cerenkov photons will have timing
determined by the group velocity \cite{group_velocity_article} in the liquid,
\begin{equation}
\label{eqGroup}
v_{g}(\lambda) = \frac{c_{vacuum}}{n(\lambda) + dn(\lambda)/dlog(\lambda)}.
\end{equation}
In Geant4 optical photons are assigned the group velocity in the wavelength 
region of normal dispersion ($dn(\lambda)/dlog(\lambda)$>0) which is relevant to 
our study. 
The longer wavelengths Cerenkov photons typically arrive before
the scintillation light, which is slowed by both the scintillation
processes and the shorter wavelengths involved. Thus, with sufficient
timing resolution and sensitivity to longer wavelength, it should be
possible to separate the directional Cerenkov light and the isotropic
scintillation light in time and then reconstruct the direction of the
initial particle.

\section{Geant4 Simulation}
In order to study the effects relevant to directionality reconstruction in liquid scintillators, a Geant4 \cite{geant4one,geant4two} simulation has been set up. The presented simulation application has been developed using standard Geant4 functionality for the optical model of the liquid scintillator. Version Geant4.9.6 is used and the simulation application has been built from a customized official application example. The simulation is kept simple to provide generally applicable information about the main factors in directionality reconstruction.

The implemented detector geometry is a sphere of 6.5~m (or 0.65~m) diameter filled with liquid scintillator. Default scintillator properties have been chosen to match the KamLAND scintillator \cite{tbd}: 80~\% n-Dodecane, 20~\% Pseudocumene (1,2,4-Trimethylbenzene) and 1.52~g/l PPO (2,5-Diphenyloxazole). The implemented scintillator properties include the atomic composition and density ($\rho$ = 0.7752~g/ml), the wavelength-dependent attenuation length and refractive index, the scintillation emission spectrum, emission rise time ($\tau_r$ = 1.0~ns) and emission decay time constants ($\tau_{d1}$ = 6.9~ns and $\tau_{d2}$ = 8.8~ns with relative weights of 0.87 and 0.13), scintillator light yield (LY, 9030.5 photons/MeV) and the Birks constant ($kB$ = 0.106~mm/MeV) \cite{tbd}. Variations from the baseline KamLAND case are discussed below when applicable. Reemission of absorbed photons in the scintillator bulk volume and scattering has not been included. Reflections, reemission and scattering are expected to be corrections to the main effects studied in this paper and are subject of future work.

The sphere surface is used as the photodetector. It is treated as fully absorbing (no reflections) with a coverage of 100~\%. Two important photodetector properties have also been varied: Transit time spread (TTS, default $\sigma$ = 0.1~ns) and wavelength-dependent quantum efficiency (QE) for photoelectron production. Default is the QE of a bialkali photocathode (Hamamatsu R7081 PMT), digitized values come from the Double Chooz \cite{dctwo} Monte Carlo simulation. The KamLAND 17 inch PMTs use the same photocathode type (bialkali) with similar quantum efficiency.

%deprecated text: fake vertex reconstruction not needed anymore. 
%To this end, we use the KamLAND vertex resolution of 12.0~cm \cite{tbd} ($\sigma$ in one dimension) to randomly draw a vertex around the known true vertex for each event. This vertex is subsequently used as a reconstructed vertex and consequently, a time of flight (TOF) correction relative to the center of the sphere is applied for each photon hit depending on the distance between the reconstructed vertex and the hit position on the sphere. For the TOF correction a single effective speed of light is used. This effective speed was extracted at the mean wavelength of both Scintillation and Cerenkov photons (408~nm). The 12.0~cm vertex resolution corresponds to a time resolution broadening of about 0.6~ns ($\sigma$). Note that the vertex resolution itself is dependent on scintillator and detector properties. For example, the PMTs used in KamLAND have a TTS of 1.28~ns ($\sigma$) and significant improvement of the vertex resolution could be possible with faster photodetectors. In section \ref{reconstruction_sec}, we discuss first results on more realistic reconstruction.
\begin{figure*}[tbh]
        \begin{center}
        \subfigure[ Default simulation.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}}
        \subfigure[ Increased TTS (1.28~ns).]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}}
        \subfigure[ Red-sensitive photocathode.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_RedSensitiveQE_KamlandScintSpec_TIME.pdf}}
        \caption[]{PE times after TTS application for the simulation of 1000 electrons (5~MeV) with different settings. PEs from Cerenkov light (black) and scintillation light (grey) are compared. The dashed vertical line illustrates a time cut at 34~ns. (a) Default simulation: bialkali photocathode and TTS = 0.1~ns ($\sigma$). The number of PEs per event after the 34.0~ns time cut is 171 from scintillation and 108 from Cerenkov. (b) Default simulation settings but for TTS = 1.28~ns (KamLAND 17 in. PMTs). The number of PEs per event after the 34.0~ns time cut is 349 from scintillation and 88 from Cerenkov. (c) Default simulation settings but for GaAsP photocathode. The number of PEs per event after the 34.0~ns time cut is 50.8 from scintillation and 171 from Cerenkov. \label{time_plots_comparison}}
        \end{center}
\end{figure*}
Before we discuss the simulation results for different simulation settings in the following sections, we highlight the effects which contribute to the timing of the scintillator detector system. First, the simulated travel time of the initial 5~MeV electron is between 0.10 and 0.15~ns, while the travel distance is about 3~cm. Second, the scintillation light emission follows a scintillator-specific distribution characterized by rise and decay time(s). Before the solutes in liquid scintillator can emit optical photons, the energy has to be transfered from the solvent to the solute. The time constant of this energy transfer accounts for a rise time in scintillation light emission. Past neutrino experiments were not highly interested in the effect of scintillation rise time which is the reason why there is a lack of accurate numbers. We assume a rise time of 1~ns, more detailed studies are needed in the future. The two time constants used to describe the falling edge of the scintillator emission time distribution (quoted above) are values specific to the KamLAND scintillator. Third, chromatic dispersion turns out to be an important effect in a 6.5~m diameter detector at the level of precision needed for direction reconstruction. Due to the wavelength-dependence of the refractive index the speed of light in the scintillator (the group velocity is used in Geant4 for normal dispersion) is also wavelength dependent. In order to study the size of this effect, we extracted results from a simplified simulation of 5~MeV electrons at the center of the sphere where we used instantaneous scintillation and did apply neither TTS nor TOF correction. The true time distributions of photoelectrons were analyzed for scintillation light and Cerenkov light separately. Photoelectrons coming from Cerenkov light are created about 0.5~ns earlier than PEs from scintillation light on average. The RMS values for both the Cerenkov and scintillation light true PE time distributions are both about 0.5~ns. Note that these numbers include the effect of the finite electron travel time.

On the detector side, the time information of single photoelectrons is affected by the TTS of the photodetectors, a number which can be different by orders of magnitude depending on the detector type. The default TTS of 0.1~ns ($\sigma$) is a value which can be achieved with the large area picosecond photodetectors (LAPPDs)\cite{LAPPDSum,LAPPDTDR} and possibly hybrid photodetectors (HPDs)\cite{hpdThesis}. In fact, even significantly lower TTS numbers are realistic with the LAPPD. Finally, the accuracy of the vertex reconstruction is reflected in broadening and distortions of the time spectrum (after the TOF correction has been applied). Another effect which is to be addressed in future work is the granularity of the photodetectors which adds to the uncertainty in the light path and thus the TOF correction.

In Sections \ref{detector_timing_sec} to \ref{edep_size_sec}, we comparatively study the photoelectron timing for different detector properties. We follow the idea to increase discrimination between Cerenkov and scintillation light by detector timing. The main quantities provided by the Geant4 simulation are the photoelectron hit positions and the detection times after TTS has been applied. In section \ref{reconstruction_sec} these quantities are used for event reconstruction.
%%statement that rise time is not super-crucial (see April 11 talk)
%% include wavelength spectra, emission spectra, absorption (?), refractive index  ?
%% Other settings: physics list, theCerenkovProcess->SetMaxNumPhotonsPerStep(20);theCerenkovProcess->SetMaxBetaChangePerStep(10.0); theCerenkovProcess->SetTrackSecondariesFirst(true); theScintillationProcess->SetScintillationYieldFactor(1.); theScintillationProcess->SetTrackSecondariesFirst(true); theScintillationProcess->SetFiniteRiseTime(true); particle gun, 5MeV e- in x direction from center, random polarization
%% discussion: can chromatic dispersion be disentangled? With very high statistics maybe yes (see Ben Monreal, Matt Wetstein).
%% discussion: many scintillator properties depend on each other. Optimization of scintillator properties could lead to better results for directionality reconstruction. i.e. slower scintillation (but vertex reco worse!, LY tends to be lower) , lower wavelength, 
%% idea: filter for part of the PMTs or two kinds of PMTs: separation by wavelength
%% coverage is just a reduction by factor 0.22 (flat over wavelengths etc.), not studied in the paper?  
%% at the 0.1 ns level in a 6.5 m detector one has to be careful with the input of nref. If the slope is not smooth enough the group velocity calcuation is creating wavelength-dependent distortions. 
%% Reflections and absorption/reemission and scattering will shift events to higher times were they will likely not pass the time cut. They will effectively change the scintillation/Cerenkov ratio. Scattering might distort the angular variables and is subject of future work. 
%% In small det. the chromatic dispersion doesn't help too much. (In the small detector the detector properties are the driving factor.)
%% discussion: some scintillator and detector properties have complicated effects on the directionality, also other requirements have to be considered (high LY, high resolutionc). Multiparameter optimization problem.  
%% nuebar scattering --> one of next paper

\section{Detector Timing}
\label{detector_timing_sec}
%%\begin{figure}
%%        \begin{center}
%%        \includegraphics[scale=0.40]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}
%%        \caption[]{PE times after TTS application for the simulation of 1000 electrons (5~MeV) with default settings: Detector diameter = 6.5~m, bialkali photocathode, KamLAND scintillator emission spectrum, TTS = 0.1~ns ($\sigma$). PEs from Cerenkov light (black) and scintillation light (red) are compared. The number of PEs per event after a 34.0~ns time cut is 171 from scintillation and 108 from Cerenkov. \label{6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME}}
%%        \end{center}
%%\end{figure}
%%\begin{figure}
%%        \begin{center}
%%        \includegraphics[scale=0.40]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}
%%        \caption[]{PE times after TTS application for the simulation of 1000 electrons (5~MeV) with TTS = 1.28~ns: Detector diameter = 6.5~m, bialkali photocathode, KamLAND scintillator emission spectrum, TTS = 1.28~ns ($\sigma$). PEs from Cerenkov light (black) and scintillation light (red) are compared. The number of PEs per event after a 34.0~ns time cut is 349 from scintillation and 88 from Cerenkov. \label{6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME}}
%%        \end{center}
%%\end{figure}

First we discuss the default simulation settings which are described in the previous section. Figure \ref{time_plot_comparison} (a) shows the TTS-smeared photoelectron detection time for 1000 simulated electrons with 5~MeV energy in the center of the detector with initial momentum direction coinciding with the x-axis. The photoelectrons induced by Cerenkov light arrive earlier, as expected due to the instantaneous emission and the differences in the photon speed. There is however overlap with the PE time distribution coming from scintillation light. In order to compare simulations with different parameters to each other a fixed time cut $t \leq$ 34.0~ns (or 33.0~ns?) is applied (using the truth information of the electron starting time) instead of an event-by-event time reconstruction (see Section \ref{reconstruction_sec} for reconstruction results). For the default simulation case, the number of PEs per event coming from Cerenkov light after the 34.0~ns time cut (108) is 98~\% of the total number of PEs from Cerenkov light (110). For scintillation the number of PEs after the time cut (171) is only 3.2~\% of the total scintillation-induced PEs (5445).  
\begin{figure}[bt]
        \begin{center}
        \includegraphics[scale=0.40]{graphs/cos_psi_comp_h.pdf}
        \caption[]{Angular distribution of photoelectron hits relative to the original electron direction for 1000 events (5~MeV electrons at the center in x direction): $\cos(\Psi) = x_{hit}/\vec{r}_{hit}$. Default simulation settings were used and both Cerenkov and Scintillation light are included. A 33.0~ns time cut (black) is compared with a 34.0~ns time cut (grey). \label{cerenkov_cone}}
        \end{center}
\end{figure}

Figure \ref{cerenkov_cone} displays the PE hit pattern after the 34.0~ns time cut has been applied (1000 events plotted together). Although this time cut is an oversimplification of actual time reconstruction effects, we can use it to indicate the spatial distribution of hits after timing information has been used to separate Cerenkov and scintillation light. The Cerenkov ring structure can be clearly seen (opening angle between 40\textdegree and 50\textdegree) which demonstrates that the directional signal conveyed by the Cerenkov photons is not erased by scattering of the initial 5~MeV electrons.

If the 17 inch KamLAND PMTs \cite{tbd} (TTS = 1.28~ns) are used in the simulation, the broadening of the time distributions leads to a strongly decreased ratio of Cerenkov over Scintillation light after the time cut (see FIG. \ref{time_plots_comparison} (b)). This demonstrates that low photodetector TTS is critical for directionality reconstruction and motivates use of novel detector types.

\section{Detector Wavelength Response}
\label{detector_wavelength_response_sec} 
Comlementary to lowering the photodetector TTS, optimization of the wavelength-dependent QE is promising. Since Cerenkov photons which passed 6.5~m of scintillator have higher average wavelengths than scintillation photons, a photodetector which is more sensitive at high wavelengths increases not only the absolute number of PEs but also the ratio between Cerenkov- and scintillation-induced PEs. Therefore, a simulation has been run with the QE of an extended red-sensitive GaAsP photocathode (digitized Hamamatsu R3809U-63 QE data was used).
%%\begin{figure}
%%        \begin{center}
%%        \includegraphics[scale=0.40]{graphs/6p5Meter_5MeVElectrons_RedSensitiveQE_KamlandScintSpec_TIME.pdf}
%%        \caption[]{Effectively reconstructed PE times for the simulation of 1000 electrons (5~MeV) with red-sensitive photocathode: Detector diameter = 6.5~m, red-sensitive GaAsP photocathode, KamLAND scintillator emission spectrum, TTS = 0.1~ns ($\sigma$). PEs from Cerenkov light (black) and scintillation light (red) are compared. The number of PEs per event after a 34.0~ns time cut is 50.8 from scintillation and 171 from Cerenkov. \label{6p5Meter_5MeVElectrons_RedSensitiveQE_KamlandScintSpec_TIME}}
%%        \end{center}
%%\end{figure}
Figure \ref{time_plots_comparison} (c) shows the results for the modified simulation with high QE in the red spectral region. The higher absolute number of PE coming from Cerenkov light (factor of $\approx$ 2) and the increased Cerenkov/scintillation ratio (factor of $\approx$ 1.6) after the time cut would significantly enhance the directionality reconstruction capabilities.

\section{Scintillator Emission Spectrum}
\label{scintillator_emission_sec}
One important contributing factor to the separation in time between Cerenkov and scintillation photon hits is the higher average light speed for the photons coming from the Cerenkov effect due to their higher average wavelength. In this section we present two more sets of simulations where the scintillator emission spectrum was changed.

Recently, the use of quantum dots (QDs) in liquid scintillators has been studied as a possibility to improve future large scale neutrino experiments \cite{tbd}. One important motivation is the control of the emission spectra by tuning the size or composition of the quantum dots. The emission spectrum of alloyed core/shell CdS$_x$Se$_{1-x}$/ZnS Trilite450 \cite{tbd} quantum dots was measured. This spectrum is a symmetric peak centered around 461~nm with FWHM = 29~nm.

%%\begin{figure}
%%        \begin{center}
%%        \includegraphics[scale=0.40]{graphs/6p5Meter_5MeVElectrons_Bialkali_QD384nmScintSpec_TIME.pdf}
%%        \caption[]{PE times for the simulation of 1000 electrons (5~MeV) with a quantum dot emission spectrum centered at 384~nm: Detector diameter = 6.5~m, bialkali photocathode, QD emission spectrum at 384~nm, TTS = 0.1~ns ($\sigma$). PEs from Cerenkov light only (black) and scintillation light only (red) are compared. The number of PEs per event after a 34.0~ns time cut is 124 from scintillation and 107 from Cerenkov. \label{6p5Meter_5MeVElectrons_Bialkali_QD384nmScintSpec_TIME}}
%%        \end{center}
%%\end{figure}

In order to isolate the effect of a different emission spectrum, the other simulation settings, including the KamLAND absorption spectrum are kept unchanged. Compared to the default case shown in FIG. \ref{time_plot_comparison} (a) the separation is worse because the scintillation light wavelengths are higher than in the KamLAND emission spectrum. However, advances in the production of commercial quantum dot samples could yield quantum dots which have similar, single peak emission shapes at lower wavelengths. This case has been simulated using the same spectral shape of the measured Trilite450 emission but shifted to lower wavelengths such that the emission peak is centered at 384~nm. This peak emission value has been measured for other types of QDs, however with a much more pronounced tail \cite{tbd}. The resulting PE time distribution shows a slightly clearer separation of Cerenkov and Scintillation light. After the 34.0~ns cut on the TTS-smeared PE time we obtain a higher Cerenkov/scintillation ratio compared to the default simulation: The number of Cerenkov-induced PE after the time cut is unchanged while the number of PEs coming from scintillation light is decreased due to the higher average photon travel times.

%%idea: use (some) PMTs with high QE for high lambda (and/or low QE for normal lambda). 
%% add some more info why we are interested in a smaller detector
%% motivation: Daedalus/ISODAR (?), 0nuBB 
\section{Reconstruction}
\label{reconstruction_sec}

@@

%%20 inch PMTs: 
%%H.~Kume, S.~Sawaki, M.~Ito, K.~Arisaka, T.~Kajita, A.~Nishimura, and A.~Suzuki, Nucl. Instrum. Methods Phys. Res. \bf{205}, 443 (1983).
%%17 inch PMTs:

>>>>

\section{Light Production in Liquid Scintillators}
Liquid scintillators are cocktails of aromatic hydrocarbons. When
charged particles move through a scintillator, the molecules are
excited, predominantly the the non-localized electrons in the
$\pi$-bonds of the phenyl groups~\cite{birks_book}. Vibrational
and rotational modes of the molecules are turned into heat within
picoseconds through collisions with other molecules.  Within $\sim$10
picoseconds, the $\pi$-electrons de-excite to the first excited state
from higher levels through radiationless transitions. The first
excited state de-excites through photon emission. There are two
characteristic times for this de-excitation, depending if the singlet
state or the triplet state was excited.  The singlet state will
de-excite within nanoseconds while the triplet state de-excites on the
order of 10's or 100's of nanoseconds. These two processes are
fluorescence and phosphorescence respectively. The exact time
constants for these processes are determined by the composition of the
scintillator.

@@

A good liquid scintillator will produce $\sim$10,000 photons
isotropically per MeV of deposited energy. Although less abundant,
Cerenkov light will be produced as well if a particle is moving faster
than the speed of light in the medium.  This light is emitted in a
cone pointed in the direction of the particles trajectory and with a
continuous spectrum weighted toward shorter wavelengths
but extending well into the red. The spectrum is described by:
\begin{equation}
\label{eqCerenkov}
\frac{dN}{d\lambda dx} = \frac{2 \pi \alpha Z^2}{\lambda^2} ( 1 - \frac{1}{\beta^2 n(\lambda)^2} )
\end{equation}
where $n(\lambda)$ is the wavelength-dependent index of refraction and
$\beta$ is the velocity of the incoming particle. The Cerenkov light
produced at wavelengths shorter than the absorption cutoff of the
scintillator will be absorbed and re-emitted as isotropic light, but
wavelengths longer than this cutoff will propagate across the
detector, retaining their directional information. The yield is
roughly 60 photons per MeV, assuming a 400~nm absorption
cutoff\cite{qdot}. These undisturbed Cerenkov photons will have timing
determined by the group velocity \cite{group_velocity_article} in the liquid,
\begin{equation}
\label{eqGroup}
v_{g}(\lambda) = \frac{c_{vacuum}}{n(\lambda) - dn(\lambda)/d\textnormal{log}(\lambda)}.
\end{equation}
In Geant4 optical photons are assigned the group velocity in the wavelength 
region of normal dispersion which is relevant to our study. 
The longer wavelengths Cerenkov photons typically arrive before
the scintillation light, which is slowed by both the scintillation
processes and the shorter wavelengths involved. Thus, with sufficient
timing resolution and sensitivity to longer wavelength, it should be
possible to separate the directional Cerenkov light and the isotropic
scintillation light in time and then reconstruct the direction of the
initial particle.

\section{Geant4 Simulation}
In order to study the effects relevant to directionality reconstruction in liquid scintillators, a Geant4 \cite{geant4one,geant4two} simulation has been set up. The presented simulation has been developed using standard Geant4 (version 4.9.6) functionality for the liquid scintillator optical model. The simulation has been built from a customized official Geant4 application example and it is kept rather generic to provide generally applicable information about the main factors in directionality reconstruction.

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.25]{graphs/detector_view.png}
        \caption[]{Two back-to-back electrons with 1.41~MeV each (equally divided energy of $^116$Cd $0\nu\beta\beta$ decay) at the center of the sphere with initial directions along the x and -x-axis. Only Cerenkov photons are drawn to illustrate the directionality of the event. \label{detector_view}}
        \end{center}
\end{figure}

The implemented detector geometry is a sphere of 6.5~m diameter filled with liquid scintillator (see Figure \ref{detector_view}). Default scintillator properties have been chosen to match the KamLAND scintillator \cite{kamland2003}: 80~\% n-Dodecane, 20~\% Pseudocumene (1,2,4-Trimethylbenzene) and 1.52~g/l PPO (2,5-Diphenyloxazole). The implemented scintillator properties include the atomic composition and density ($\rho$ = 0.78~g/ml), the wavelength-dependent attenuation length \cite{tajimaMaster} and refractive index \cite{OlegThesis}, the scintillation emission spectrum, emission rise time ($\tau_r$ = 1.0~ns) and emission decay time constants ($\tau_{d1}$ = 6.9~ns and $\tau_{d2}$ = 8.8~ns with relative weights of 0.87 and 0.13) \cite{tajimaThesis}, scintillator light yield (LY, 9030 photons/MeV) and the Birks constant ($kB$ $\approx$ 0.1~mm/MeV) \cite{ChrisThesis}. Variations from the baseline KamLAND case are discussed below when applicable. Reflections, reemission of absorbed photons in the scintillator bulk volume and scattering has not been included. These effects are expected to be corrections to the main effects studied in this paper and are subject of future work.

The inner sphere surface is used as the photodetector. It is treated as fully absorbing (no reflections) with a photodetector coverage of 100~\%. Two important photodetector properties have been varied: Transit time spread (TTS, default $\sigma$ = 0.1~ns) and wavelength-dependent quantum efficiency (QE) for photoelectron production. Default is the QE of a bialkali photocathode (Hamamatsu R7081 PMT), digitized values come from the Double Chooz \cite{dctwo} Monte Carlo simulation. We note that the KamLAND 17 inch PMTs use the same photocathode type (bialkali) with similar quantum efficiency.

\begin{figure*}[tbh]
\begin{center}
        \subfigure[ ~Default simulation.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_TIME.pdf}}
        \subfigure[ ~Increased TTS (1.28~ns).]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_Bialkali_KamlandScintSpec_1p28nsTTS_TIME.pdf}}
        \subfigure[ ~Red-sensitive photocathode.]{\includegraphics[scale=0.295]{graphs/6p5Meter_5MeVElectrons_RedSensitiveQE_KamlandScintSpec_TIME.pdf}}
        \caption[]{PE times after TTS application for the simulation of 1000 electrons (5~MeV) with different settings. PEs from Cerenkov light (black) and scintillation light (grey) are compared. The dashed vertical line illustrates a time cut at 34~ns. (a) Default simulation: bialkali photocathode and TTS = 0.1~ns ($\sigma$). After the 34.0~ns time cut we get 171~PEs from scintillation and 108~PEs from Cerenkov light. (b) Default simulation settings except for TTS = 1.28~ns (KamLAND 17 in. PMTs). After the 34.0~ns time cut we get 349~PEs from scintillation and 88~PEs from Cerenkov light. (c) Default simulation settings except for a GaAsP photocathode. After the 34.0~ns time cut we get 226~PEs from scintillation and 229~PEs from Cerenkov light. \label{time_plots_comparison}}
        \end{center}
\end{figure*}
Before the simulation results for different simulation settings are discussed in the following sections, we highlight the effects which contribute to the timing of the scintillator detector system. First, the simulated travel time of the initial 5~MeV electron is between 0.10 and 0.15~ns, while the travel distance is about 3~cm. Second, the scintillation light emission follows a scintillator-specific distribution characterized by rise and decay time(s). Before the solutes in liquid scintillator can emit optical photons, the energy has to be transfered from the solvent to the solute. The time constant of this energy transfer accounts for a rise time in scintillation light emission. Past neutrino experiments were not highly interested in the effect of scintillation rise time which is the reason why there is a lack of accurate numbers. We assume a rise time of 1.0~ns, more detailed studies are needed in the future. The two time constants used to describe the falling edge of the scintillator emission time distribution (quoted above) are values specific to the KamLAND scintillator. Third, chromatic dispersion turns out to be an important effect in a 6.5~m diameter detector at the level of precision needed for direction reconstruction. Due to the wavelength-dependence of the refractive index the speed of light in the scintillator (see Equation (\ref{eqGroup})) increases with increasing photon wavelengths for normal dispersion. In order to study the size of this effect, we extracted results from a simplified simulation of 5~MeV electrons at the center of the sphere where we used instantaneous scintillation and applied QE but no TTS. The true hit time distributions of photoelectrons were analyzed for scintillation light and Cerenkov light separately. Photoelectrons coming from Cerenkov light are on average created about 0.5~ns earlier than PEs from scintillation light. The RMS values for both the Cerenkov and scintillation light true PE time distributions are both about 0.5~ns. Note that these numbers include the effect of the finite electron travel time.

On the detector side, the time information of single photoelectrons is affected by the TTS of the photodetectors, a number which can be different by orders of magnitude depending on the detector type. The default TTS of 0.1~ns ($\sigma$) is a value which can be achieved with the large area picosecond photodetectors (LAPPDs)\cite{LAPPDSum,LAPPDTDR} and possibly hybrid photodetectors (HPDs)\cite{hpdThesis}. In fact, even significantly lower TTS numbers are realistic with the LAPPD. Finally, the accuracy of the vertex reconstruction is reflected in broadening and distortions of the time spectrum (after time-of-flight correction has been applied). Another effect which is to be addressed in future work is the granularity of the photodetectors which adds to the uncertainty in the light path and thus the time-of-flight correction.

In Sections \ref{detector_timing_sec} to \ref{scintillator_emission_sec}, we comparatively study the photoelectron timing for different detector configurations. We follow the idea to increase discrimination between Cerenkov and scintillation light by accurate detector timing. The main relevant quantities provided by the Geant4 simulation are the photoelectron hit positions and the detection times after TTS has been applied. In section \ref{reconstruction_sec} these quantities are used for event reconstruction.

\section{Detector Timing}
\label{detector_timing_sec}

First we discuss results for the default simulation settings which are described in the previous section. Figure \ref{time_plots_comparison} (a) shows the TTS-smeared photoelectron detection times for 1000 simulated electrons with 5~MeV energy in the center of the detector with initial momentum direction coinciding with the x-axis. The photoelectrons induced by Cerenkov light arrive earlier, as expected due to the instantaneous emission and the higher average photon speed (higher wavelengths) compared to scintillation after QE weighting. There is however significant overlap with the PE time distribution coming from scintillation light. In order to compare simulations with different parameters to each other a fixed time cut $t \leq$ 34.0~ns (or 33.0~ns) is applied using the truth information of the electron starting time instead of an event-by-event time reconstruction (see Section \ref{reconstruction_sec} for reconstruction results). For the default simulation case, the number of PEs per event coming from Cerenkov light after the 34.0~ns time cut (108) is 98~\% of the total number of PEs from Cerenkov light (110). For scintillation the number of PEs after the time cut (171) is only 3.2~\% of the total scintillation-induced PEs (5445). This indicates the effectiveness of a time cut to separate Cerenkov light from scintillation light. An important figure of merit when comparing different simulation settings is the ratio of Cerenkov/scintillation induced PE after the 34.0~ns time cut since a higher ratio means more directional information per PE. The ratio is 0.63 for the default settings. Note that this ratio depends on the time cut, which will be optimized in future work together with the absolute number of PEs per event passing the cut. 
\begin{figure}
        \begin{center}
        \includegraphics[scale=0.40]{graphs/cos_psi_comp_h.pdf}
        \caption[]{Angular distribution of photoelectron hits relative to the original electron direction along the x-axis: $\cos(\Psi) = x_{hit}/\vec{r}_{hit}$. 1000 electron events with 5~MeV at the center are summed. Default simulation settings are used and both Cerenkov and scintillation light are included. A 33.0~ns time cut (black) is compared with a 34.0~ns time cut (grey). \label{cerenkov_cone}}
        \end{center}
\end{figure}

Figure \ref{cerenkov_cone} displays the PE hit pattern after cutting on time. Although this time cut is a simplification of actual time reconstruction effects, we can use it to indicate the spatial distribution of hits after timing information has been used to separate Cerenkov and scintillation light. The Cerenkov ring structure can be clearly seen (peak at about 46\textdegree) which demonstrates that the directional signal conveyed by the Cerenkov photons is not erased by scattering of the initial 5~MeV electrons.

If the 17 inch KamLAND PMTs \cite{tajimaMaster,kume_1983} (TTS = 1.28~ns) are used in the simulation, the broadening of the time distributions leads to a strongly decreased ratio of Cerenkov over scintillation light (0.25) after the 34~ns time cut (see FIG. \ref{time_plots_comparison} (b)). This shows that low photodetector TTS is critical for directionality reconstruction and motivates the use of novel detector types.

\section{Detector Wavelength Response}
\label{detector_wavelength_response_sec} 
Complementary to lowering the photodetector TTS, optimization of the wavelength-dependent QE is promising. Since Cerenkov photons which passed 6.5~m of scintillator have higher average wavelengths than scintillation photons, a photodetector which is more sensitive at high wavelengths increases not only the absolute number of PEs but also the ratio between Cerenkov- and scintillation-induced PEs. Therefore, a simulation has been run with the QE of an extended red-sensitive GaAsP photocathode (Hamamatsu R3809U-63 QE data was used).

Figure \ref{time_plots_comparison} (c) shows the results for the modified simulation with high QE in the red spectral region. The higher absolute number of PE coming from Cerenkov light (factor of $\approx$ 2) and the increased Cerenkov/scintillation ratio (1.6) after the time cut would significantly enhance the directionality reconstruction capabilities given that high coverage can be achieved with red-sensitive photocathodes.

\section{Scintillator Emission Spectrum}
\label{scintillator_emission_sec}
An alternative route towards increasing the separation in time between Cerenkov and scintillation photon hits is the tuning of the scintillator emission spectrum. Recently, the use of quantum dots (QDs) in liquid scintillators has been studied as a possibility to improve future large scale neutrino experiments \cite{qdot2}. One major motivation for quantum-dot-doped scintillator is control of the emission spectra by tuning the size or composition of the quantum dots. The emission spectrum of commercial alloyed core/shell CdS$_x$Se$_{1-x}$/ZnS quantum dots was measured. This spectrum shows a symmetric peak centered around 461~nm with FWHM = 29~nm.

In order to isolate the effect of a different emission spectrum, the other simulation settings, including the KamLAND absorption spectrum are kept unchanged. Compared to the default case shown in FIG. \ref{time_plots_comparison} (a) the separation is worse (as expected) because the scintillation light wavelengths are higher than in the KamLAND emission spectrum. However, advances in the production of commercial quantum dot samples could yield quantum dots which have similar, single peak emission shapes at lower wavelengths. This case has been simulated using the same spectral shape of the measured core-shell quantum dot emission but shifted to lower wavelengths such that the emission peak is centered at 384~nm. This peak emission value has been measured for other types of QDs, however with a much more pronounced tail \cite{qdot2}. The resulting PE time distribution shows improved separation of Cerenkov and scintillation light compared to the default simulation. After the 34.0~ns cut on the TTS-smeared PE time we obtain a higher Cerenkov/scintillation ratio of 0.86 (107 PE from Cerenkov light and 124 PE from scintillation). The number of Cerenkov-induced PE after the time cut is unchanged while the number of PEs coming from scintillation light is decreased due to the higher average photon travel times.

\section{Reconstruction}
\label{reconstruction_sec}

@@
